---
title: "detect Mutations - Swiss data"
author: Sylvain Schmitt
date: April 20, 2021
output:
  github_document: 
    toc: yes
    toc_depth: 2
---

```{r setup, include=FALSE}
rm(list = ls()) ; invisible(gc()) ; set.seed(42)
library(knitr)
library(kableExtra)
if(knitr:::is_html_output()) options(knitr.table.format = "html") 
if(knitr:::is_latex_output()) options(knitr.table.format = "latex") 
library(tidyverse)
library(vcfR)
library(Biostrings)
theme_set(bayesplot::theme_default())
opts_chunk$set(echo = F, message = F, warning = F, fig.height = 6, fig.width = 8,
               cache = F, cache.lazy = F)
```

[`singularity` & `snakemake`](https://github.com/sylvainschmitt/snakemake_singularity) workflow to detect mutations with several alignment and mutation detection tools.

```{r dag}
knitr::include_graphics("dag/dag.minimal.svg")
```

# Installation

- [x] Python ≥3.5
- [x] Snakemake ≥5.24.1
- [x] Golang ≥1.15.2
- [x] Singularity ≥3.7.3
- [x] This workflow

```{bash, eval=F, echo=T}
# Python
sudo apt-get install python3.5
# Snakemake
sudo apt install snakemake`
# Golang
export VERSION=1.15.8 OS=linux ARCH=amd64  # change this as you need
wget -O /tmp/go${VERSION}.${OS}-${ARCH}.tar.gz https://dl.google.com/go/go${VERSION}.${OS}-${ARCH}.tar.gz && \
sudo tar -C /usr/local -xzf /tmp/go${VERSION}.${OS}-${ARCH}.tar.gz
echo 'export GOPATH=${HOME}/go' >> ~/.bashrc && \
echo 'export PATH=/usr/local/go/bin:${PATH}:${GOPATH}/bin' >> ~/.bashrc && \
source ~/.bashrc
# Singularity
mkdir -p ${GOPATH}/src/github.com/sylabs && \
  cd ${GOPATH}/src/github.com/sylabs && \
  git clone https://github.com/sylabs/singularity.git && \
  cd singularity
git checkout v3.7.3
cd ${GOPATH}/src/github.com/sylabs/singularity && \
  ./mconfig && \
  cd ./builddir && \
  make && \
  sudo make install
# detect Mutations
git clone git@github.com:sylvainschmitt/detectMutations.git
cd detectMutations
```

# Usage

## Get data

* Reference genome 3P from http://urgi.versailles.inra.fr/download/oak
* Raw reads from ENA project PRJEB8388
* Napoleon's genome (https://www.ebi.ac.uk/ena/browser/api/fasta/GCA_003013145.1?download=true&gzip=true)
* Napoleon's original mutations (https://docs.google.com/spreadsheets/d/10SmTGXCfQKU6M8KLqF_lzHDFM4A64pdpVUeNdI7MapA/edit#gid=0)

```{bash, eval=F, echo=T}
cd data/swiss
mkdir out
sarray -J swissData -o out/%j.out -e out/%j.err -t 1:00:00 --mail-type=ALL  get_reads.sh
sarray -J swissData -o out/%j.out -e out/%j.err -t 1:00:00 --mail-type=ALL  merge_reads.sh
sarray -J swissData -o out/%j.out -e out/%j.err -t 1:00:00 --mail-type=ALL  compress_reads.sh
rm -rf out
rm SRR*
sh get_genome.sh 
```

## Locally

```{bash, eval=F, echo=T}
snakemake -np -j 3 --resources mem_mb=10000 # dry run
snakemake --dag | dot -Tsvg > dag/dag.svg # dag
snakemake --use-singularity -j 3 --resources mem_mb=10000 # run
```

## HPC

```{bash, eval=F, echo=T}
module load bioinfo/snakemake-5.25.0 # for test on node
snakemake -np # dry run
sbatch job.sh # run
snakemake --dag | dot -Tsvg > dag/dag.svg # dag
```

# Workflow

## Reference

*Copy and index reference for software to work with.*

### [samtools_faidx_split](https://github.com/sylvainschmitt/detectMutations/blob/swiss/rules/samtools_faidx_split.smk)

* Tools: [`samtools faidx`](http://www.htslib.org/doc/samtools-faidx.html)
* Singularity: oras://registry.forgemia.inra.fr/gafl/singularity/samtools/samtools:latest

### [bwa_index](https://github.com/sylvainschmitt/detectMutations/blob/swiss/rules/bwa_index.smk)

* Tools: [`BWA index`](http://bio-bwa.sourceforge.net/bwa.shtml)
* Singularity: oras://registry.forgemia.inra.fr/gafl/singularity/bwa/bwa:latest

### [samtools_faidx](https://github.com/sylvainschmitt/detectMutations/blob/swiss/rules/samtools_faidx.smk)

* Tools: [`samtools faidx`](http://www.htslib.org/doc/samtools-faidx.html)
* Singularity: oras://registry.forgemia.inra.fr/gafl/singularity/samtools/samtools:latest

### [gatk_dict](https://github.com/sylvainschmitt/detectMutations/blob/swiss/rules/gatk_dict.smk)

* Tools: [`gatk CreateSequenceDictionary`](https://gatk.broadinstitute.org/hc/en-us/articles/360036729911-CreateSequenceDictionary-Picard-)
* Singularity: docker://broadinstitute/gatk

## Napoleon

*Get the position of original mutations from Napoleon on the 3P genome.*

### [cp_napo](https://github.com/sylvainschmitt/detectMutations/blob/swiss/rules/cp_napo.smk)

* Tools: `cp`

### [samtools_faidx_napo](https://github.com/sylvainschmitt/detectMutations/blob/swiss/rules/samtools_faidx_napo.smk)

* Tools: [`samtools faidx`](http://www.htslib.org/doc/samtools-faidx.html)
* Singularity: oras://registry.forgemia.inra.fr/gafl/singularity/samtools/samtools:latest

### [napomutations2bed](https://github.com/sylvainschmitt/detectMutations/blob/swiss/rules/napomutations2bed.smk)

* Script: [`napomutations2bed.R`](https://github.com/sylvainschmitt/detectMutations/blob/swiss/scripts/napomutations2bed.R)
* Singularity: to be added, currently uses local install

### [bedtools_getfasta](https://github.com/sylvainschmitt/detectMutations/blob/swiss/rules/bedtools_getfasta.smk)

* Tools: [`bedtools getfasta`](https://bedtools.readthedocs.io/en/latest/content/tools/getfasta.html)
* Singularity: oras://registry.forgemia.inra.fr/gafl/singularity/bedtools/bedtools:latest

### [blat](https://github.com/sylvainschmitt/detectMutations/blob/swiss/rules/blat.smk)

* Tools: [`blat](http://genome.ucsc.edu/FAQ/FAQblat.html)
* Singularity: to be added, currently uses local install

### [psl2pos](https://github.com/sylvainschmitt/detectMutations/blob/swiss/rules/psl2pos.smk)

* Script: [`psl2pos.R`](https://github.com/sylvainschmitt/detectMutations/blob/swiss/scripts/psl2pos.R)
* Singularity: to be added, currently uses local install

## Reads

*Copy reads, report quality and trim.*

### [cp_reads](https://github.com/sylvainschmitt/detectMutations/blob/swiss/rules/cp_reads.smk)

* Tools: `cp`

### [trimmomatic](https://github.com/sylvainschmitt/detectMutations/blob/swiss/rules/trimmomatic.smk)

* Tools: [`Trimmomatic`](http://www.usadellab.org/cms/uploads/supplementary/Trimmomatic/TrimmomaticManual_V0.32.pdf)
* Singularity: oras://registry.forgemia.inra.fr/gafl/singularity/trimmomatic/trimmomatic:latest

## Alignments

*Align reads against reference, mark duplicated, and report alignment quality.*

### [bwa_mem](https://github.com/sylvainschmitt/detectMutations/blob/swiss/rules/bwa_mem.smk)

* Tools: [`BWA mem`](http://bio-bwa.sourceforge.net/bwa.shtml)
* Singularity: oras://registry.forgemia.inra.fr/gafl/singularity/bwa/bwa:latest

### [samtools_view](https://github.com/sylvainschmitt/detectMutations/blob/swiss/rules/samtools_view.smk)

* Tools: [`Samtools view`](http://www.htslib.org/doc/samtools-view.html)
* Singularity: oras://registry.forgemia.inra.fr/gafl/singularity/samtools/samtools:latest

### [samtools_sort](https://github.com/sylvainschmitt/detectMutations/blob/swiss/rules/samtools_sort.smk)

* Tools: [`Samtools sort`](http://www.htslib.org/doc/samtools-sort.html)
* Singularity: oras://registry.forgemia.inra.fr/gafl/singularity/samtools/samtools:latest

### [samtools_index](https://github.com/sylvainschmitt/detectMutations/blob/swiss/rules/samtools_index.smk)

* Tools: [`Samtools index`](http://www.htslib.org/doc/samtools-index.html)
* Singularity: oras://registry.forgemia.inra.fr/gafl/singularity/samtools/samtools:latest

### [gatk_markduplicates](https://github.com/sylvainschmitt/detectMutations/blob/swiss/rules/gatk_markduplicates.smk)

* Tools: [`gatk MarkDuplicates`](https://gatk.broadinstitute.org/hc/en-us/articles/360037052812-MarkDuplicates-Picard-)
* Singularity: docker://broadinstitute/gatk

### [samtools_view_md](https://github.com/sylvainschmitt/detectMutations/blob/swiss/rules/samtools_view_md.smk)

* Tools: [`Samtools view`](http://www.htslib.org/doc/samtools-view.html)
* Singularity: oras://registry.forgemia.inra.fr/gafl/singularity/samtools/samtools:latest

### [samtools_index_md](https://github.com/sylvainschmitt/detectMutations/blob/swiss/rules/samtools_index_md.smk)

* Tools: [`Samtools index`](http://www.htslib.org/doc/samtools-index.html)
* Singularity: oras://registry.forgemia.inra.fr/gafl/singularity/samtools/samtools:latest

## Mutations

### [strelka2](https://github.com/sylvainschmitt/detectMutations/blob/swiss/rules/strelka2.smk)

* Tools: [`Strelka2`](https://github.com/Illumina/strelka)
* Singularity: docker://quay.io/wtsicgp/strelka2-manta

### [strelka2tsv](https://github.com/sylvainschmitt/detectMutations/blob/swiss/rules/strelka2tsv.smk)

* Script: [`strelka2tsv.R`](https://github.com/sylvainschmitt/detectMutations/blob/swiss/scripts/strelka2tsv.R)
* Singularity: to be added, currently uses local install

### [strelkatsv2sql](https://github.com/sylvainschmitt/detectMutations/blob/swiss/rules/strelkatsv2sql.smk)

* Script: [`strelkatsv2sql.R`](https://github.com/sylvainschmitt/detectMutations/blob/swiss/scripts/strelkatsv2sql.R)
* Singularity: to be added, currently uses local install

### [gatk_haplotypecaller](https://github.com/sylvainschmitt/detectMutations/blob/swiss/rules/gatk_haplotypecaller.smk)

* Tools: [`gatk HaplotypeCaller`](https://gatk.broadinstitute.org/hc/en-us/articles/360037225632-HaplotypeCaller)
* Singularity: docker://broadinstitute/gatk

### [gatk_genotypegvcfs](https://github.com/sylvainschmitt/detectMutations/blob/swiss/rules/gatk_genotypegvcfs.smk)

* Tools: [`gatk GenotypeGVCFs`](https://gatk.broadinstitute.org/hc/en-us/articles/360037057852-GenotypeGVCFs)
* Singularity: docker://broadinstitute/gatk
