---
title: A roadmap for plant mutation detection tools with variable sequencing depth and allele frequency using two reproducible pipelines
date: '`r Sys.Date()`'
output:
  bookdown::html_document2:
    number_sections: false  
    toc: true
    toc_float: yes
  bookdown::word_document2:
    reference_docx: ./template/template.docx
  bookdown::pdf_document2:
    includes:
      before_body: ./template/doc_prefix.tex
      in_header: ./template/preamble.tex
    number_sections: false
    toc: false
linestretch: 1.5
csl: /home/sylvain/Documents/Bibliography/csl/molecol.csl
bibliography: /home/sylvain/Documents/Bibliography/library.bib
link-citations: yes
---

**Running title: plant mutation detection tools**

Authors (**alphabetic**):

* Myriam Heuertz
    - heuertzm@gmail.com
    - orcid.org/0000-0002-6322-3645
    - Univ. Bordeaux, INRAE, BIOGECO, 69 route d’Arcachon, 33610 Cestas France
* Thibault Leroy
    - orcid.org/0000-0003-2259-9723
    - Department of Botany and Biodiversity Research, University of Vienna, Rennweg 14, A-1030 Vienna, Austria
* Sylvain Schmitt
    - Corresponding author
    - sylvain.m.schmitt@gmail.com
    - orcid.org/0000-0001-7759-7106
    - CNRS, UMR EcoFoG (Agroparistech, Cirad, INRAE, Université des Antilles, Université de la Guyane), Campus Agronomique, 97310 Kourou, French Guiana
* Niklas Tysklind
    - Niklas.Tysklind@ecofog.gf
    - INRAE, UMR EcoFoG (Agroparistech, CNRS, Cirad, Université des Antilles, Université de la Guyane), Campus Agronomique, 97310 Kourou, French Guiana

\newpage 

# Abstract

*Please provide an abstract of no more than 250 words containing the major keywords.*

\newpage 

# Introduction 

# Material and Methods

## Generation of mutations

## Detection of mutations

## Virtual experiment

## Oak data reanalyses

# Results

# Discussion

# Acknowledgements

# References

<div id="refs"></div>

# Authors’ contributions

All authors conceived the ideas;
SS developed the pipelines;
SS and TL realized the virtual experiment;
SS and TL  analyzed outputs; 
SS and TL led the writing of the manuscript. 
All authors contributed critically to the drafts and gave final approval for publication.

# Data Accessibility

`generateMutations` and `detectMutations` pipelines are available on GitHub (https://github.com/sylvainschmitt/generateMutations and https://github.com/sylvainschmitt/detectMutations).
Reanalyzed reads and corresponding genomes were extracted from GenBank under accession BioProject PRJNA327502 
and from European Nucleotide Archive under project accession code PRJEB8388.

# Tables and Figures


```{r config, message=FALSE, warning=FALSE, include=FALSE}
rm(list = ls()); invisible(gc())
library(knitr)
library(tidyverse)
library(bayesplot)
theme_set(bayesplot::theme_default())
opts_chunk$set(echo = F, message = F, warning = F, 
               fig.height = 8, fig.width = 8,
               cache = T, cache.lazy = F)
stats <- read_tsv("stats.experiment4.tsv") %>% 
  group_by(caller, N, R, AF, C) %>% 
  summarise_at(c("FN", "FP", "TP", "Precision", "Recall"), mean, na.rm = T) %>% 
  mutate(caller = recode(caller,    
                         "mutect2" = "Mutect2", 
                         "freebayes" = "freebayes", 
                         "gatk" = "GATK", 
                         "strelka2" = "Strelka2", 
                         "somaticsniper" = "Somatic Sniper", 
                         "muse" = "MuSE",
                         "varscan" = "VarScan"
  ))
```


```{r fig1, fig.height=8}
stats %>% 
  # mutate(Precision = ifelse(is.na(Precision), 0, Precision)) %>% 
  reshape2::melt(c("caller", "N", "R", "AF", "C"),
                 variable.name = "metric", value.name = "metric.value") %>% 
  filter(metric %in% c("Precision", "Recall")) %>% 
  mutate(metric = factor(metric, level = c("Recall", "Precision"))) %>% 
  ggplot(aes(AF, C, fill = metric.value)) +
  geom_tile() +
  facet_grid(caller ~ metric) +
  viridis::scale_fill_viridis("", option = "inferno") +
  xlab("Allele Fraction") + ylab("Coverage") +
  theme(strip.text.y = element_text(angle = 0),
        legend.position = "bottom",
        legend.key.width = unit(40, "points"))
```


```{r fig2, fig.width=8}
stats %>%
  filter(Precision > 0.9) %>% 
  group_by(AF, C) %>% 
  filter(Recall == max(Recall)) %>% 
  filter(Precision == max(Precision)) %>% 
  group_by(AF, C, Precision, Recall) %>% 
  summarise(caller = paste(unique(caller), sep = ", ", collapse = ", ")) %>% 
  ggplot(aes(AF, C)) +
  geom_tile(aes(fill = caller, alpha = Recall)) +
  geom_text(aes(label = round(Recall, 2)), col = "black", size	= 3) +
  scale_fill_brewer("", palette = "Paired") +
  theme(legend.position = "bottom") +
  xlab("Allele Fraction") + ylab("Coverage") +
  theme(strip.text.y = element_text(angle = 0),
        legend.position = "bottom") +
  guides(fill = guide_legend(nrow = 4)) +
  scale_alpha_continuous(guide = "none")
```

```{r fig3, fig.width=8}
stats %>%
  mutate(Precision = round(TP/(TP+FP), 4), Recall = round(TP/(TP+FN), 4)) %>% 
  filter(Precision > 0.9) %>% 
  group_by(AF, C) %>% 
  filter(Recall == max(Recall)) %>% 
  filter(Precision == max(Precision)) %>% 
  group_by(AF, C, Precision, Recall) %>% 
  summarise(caller = paste(unique(caller), sep = ", ", collapse = ", ")) %>% 
  ggplot(aes(AF, C)) +
  geom_tile(aes(fill = caller, alpha = Recall)) +
  geom_text(aes(label = round(Recall, 3)), col = "black", size	= 3) +
  scale_fill_brewer("", palette = "Accent") +
  theme(legend.position = "bottom") +
  xlab("Allele Fraction") + ylab("Coverage") +
  theme(strip.text.y = element_text(angle = 0),
        legend.position = "bottom") +
  scale_alpha_continuous(guide = "none")
```

